### 2023-01-27 paper-thesis/chapters/11-variable-capture.tex
Cleanup work
Improve the variable capture chapter

The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture. Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens \cite{tapl}. This section will explore several ways of solving this.

An example term where this problem occurs is the following: What is the type of this expression (a polymorphic identity function)?
\begin{lstlisting}
\T : Type. \T : T. T
\end{lstlisting}
The algorithm so far would tell you it is \verb|T : Type -> T : T -> T|. Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|. However, the correct answer would be \verb|T : Type -> x : T -> T|. There is no way of expressing this type without renaming a variable.

@@

Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables. Using de Bruijn indices in statix is possible, but sacrifices a lot. It would require a transformation on the AST before typechecking. Modifying the AST like this causes problems, because it changes AST nodes. All editor services that rely on \verb|.ref| annotations, such as renaming, can no longer be used. It also loses a lot of the benefits of using Spoofax, since using scope graphs relies on using names.

@@

The first solution that was attempted was having a pre-analysis transformation that gives each variable a unique name. This doesn't work for a variety of reasons. The simplest being, it doesn't actually solve the problem. Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name. Furthermore, this is a pre-analysis transformation, so similarly to using de Bruijn indices, it breaks editor services such as renaming.

@@

Anytime that we introduce a new name in a type, we could check if the name already exists in the environment, and if it does, choose a different unique name. This approach is possible but tedious to implement in Statix. It requires a new relation to traverse through the type and rename. It also increases the time complexity, as constant traversals of the type are needed.

@@

The solution we found to work best in the end is to change the definition of \verb|ID|. To be precise, at the grammar level we have two sorts, \verb|RID| is a "Raw ID", being just a string. \verb|ID| will have two constructors, one being \verb|Syn|, a syntactical \verb|ID|. The second one is \verb|ScopedName|, it is defined in statix, so there is no syntax for it. It will be generated by \verb|typeOfExpr|.

@@

The \verb|ScopedName| constructor has a scope and a raw ID. The scope is used to uniquely identify the name. The main idea is that whenever we encounter a syntactical name, we replace it with a scoped name, so it unambiguous. The scope graph will never have a syntactical name in it. However, when querying the scope graph for a syntactical name, we return the lexically closest name.

@@

\begin{lstlisting}
typeOfExpr(s1, FnConstruct(Syn("T"), Var(Syn("T")), Var(Syn("T"))))

(ScopedName(s1, T) : Type)
	       [s1]
\end{lstlisting}

@@

\begin{lstlisting}
typeOfExpr(s2, Var(Syn("T")))

(ScopedName(s1, T) : Type)     (ScopedName(s2, T) : Var(ScopedName(s1, T)))
	       [s1]<----------------------------------[s2]
\end{lstlisting}

@@

Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked named. For example, the above would be transformed to:

@@

\begin{enumerate}
	\item When we encounter a \verb|ScopedName| for the first time, we keep adding ticks to the name until we find a name that has not been used before. 
	\item We define a dynamic rule \verb|Rename :: string -> string| and we store the new name we generated using this rule.
	\item When we encounter a \verb|ScopedName|, use the \verb|Rename| rule to find what the name was transformed to.
\end{enumerate}

>>>>

The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture. Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens~\cite{tapl}. This chapter will explore several ways of solving this.

An example term where this problem occurs is the following: What is the type of this expression (a polymorphic identity function)?
\begin{lstlisting}
\T : Type. \T : T. T
\end{lstlisting}
The implementation so far would tell you it is \verb|T : Type -> T : T -> T|. Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|. However, the correct answer would be \verb|T : Type -> x : T -> T|. There is no way of expressing this type without renaming a variable.

@@

Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables. Using de Bruijn indices in statix is possible, but sacrifices a lot. Many editor services (renaming, go to definition) require \verb|.ref| annotations to be set on names, and this is not possible if the names are no longer a part of the AST.

@@

The first solution that was attempted was having a pre-analysis transformation that gives each variable a unique name. This doesn't work for a variety of reasons. First of all, it doesn't actually solve the problem. Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name. Furthermore, this changes the names in the AST, so it breaks editor services such as renaming.

@@

Anytime that we introduce a new name in a type, we could check if the name already exists in the environment, and if it does, choose a different unique name. This approach is possible but tedious to implement in Statix. It requires a new relation to traverse through the type and rename. It also increases the time complexity, as many traversals of the type are needed. This was successfully implemented, but we chose against using it in favor of the solution presented in the next section.

@@

The solution we found to work best in the end is to change the definition of \verb|ID|. To be precise, at the grammar level we have two sorts, \verb|RID| is a "Raw ID", being just a string. \verb|ID| will have two constructors, one being \verb|Syn|, a syntactical \verb|ID|, referring to the lexically closest match. The second one is \verb|ScopedName|, which is defined below.

@@

The \verb|ScopedName| constructor has a scope and a raw ID. The scope is used to uniquely identify the name. The main idea is that whenever we encounter a syntactical name during type checking, we replace it with a scoped name, so it unambiguous. The scope graph will never have a syntactical name in it. However, when querying the scope graph for a syntactical name, we return the lexically closest name.

@@

\begin{lstlisting}
typeOfExpr(s1, FnConstruct(Syn("T"), Var(Syn("T")), Var(Syn("T"))))
\end{lstlisting}
\begin{tikzpicture}[scopegraph, node distance = 2.5em and 3.0em]
	\node[scope] (s1) {$s1$};
	\node[type, above of = s1] (s1l) {$\texttt{ScopedName(s1, T) : Type()}$};
	\draw (s1) edge[type] (s1l);
\end{tikzpicture}

@@

\begin{lstlisting}
typeOfExpr(s2, Var(Syn("T")))
\end{lstlisting}
\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
	\node[scope] (s1) {$s1$};
	\node[type, above of = s1] (s1l) {$\texttt{ScopedName(s1, T) : Type()}$};
	\draw (s1) edge[type] (s1l);

	\node[scope, right of = s1, xshift=18em] (s2) {$s2$};
	\node[type, above of = s2] (s2l) {$\texttt{ScopedName(s2, T) : Var(ScopedName(s1, T))}$};
	\draw (s2) edge[type] (s2l);

@@

Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked names. For example, the above would be transformed to:

@@

\begin{enumerate}
	\item When we encounter a \verb|ScopedName| in a definition, we keep adding ticks to the name until we find a name that has not been used before. 
	\item We define a dynamic rule \verb|Rename :: string -> string| and we store the new name we generated using this rule.
	\item When we encounter a \verb|ScopedName| in a variable, use the \verb|Rename| rule to find what the name was transformed to.
\end{enumerate}
